\iffalse
{\footnotesize\em This page is optional.
It's in the table of contents and it's labeled 'ACKNOWLEDGMENTS' even
though the spelling `ACKNOWLEDGEMENTS' is also correct.
This page should be at the end of the preface, if one exists, or a
separate page, if no preface is used.}
\fi

\iffalse
Recognize faculty, friends, family but not the typist, editor or persons
paid to produce the manuscript from draft copy. Recognize funding,
special permission to reproduce figures and copyrighted material.
\fi

	It is possible for me to most gratefully acknowledge only a fraction of the people who have helped me on this journey.
	
	I must include my committee: Richard Fowles who has been a mentor and friend with a great sense of humor and support through now many years. William Carlisle who has added perspective on history for me. Rudi Von Arnim whose macro modeling expertise will continue to be invaluable. Timothy Garrett, an early ``fellow traveler'' in understanding the importance of energy inputs to economic systems based on his knowledge as an atmospheric scientist. Myron Wyn Evans, an amazing scientist whose work deeply inspired me to pursue these studies. John Watkins who has added very valuable insights into my often difficult topic choices. Thomas Maloney---Tom graciously agreed to join my committee after the untimely death of Stephen Reynolds. And, Steve Reynolds my goodness we all miss you terribly. Always ready to listen and offer feedback and dig out other things to read from his capacious knowledge and library.
	
	I have been greatly influenced by the work of Peter Temin; his careful application of basic economic principles to any history with even a few scraps of available data has been an inspiration.
	
	I am grateful to Roger Fouquet who generously and early gave me access to the energy consumption data I have used in studying the English Industrial Revolution.
	
	Finally, the entire Department of Economics at the University of Utah, a truly unique department. I have learned so much and benefited enormously from my associations with the faculty, staff, and graduate students.